The Iterative Signature Algorithm for Gene Expression data

### 2009-07-15 eisa/inst/doc/EISA_tutorial.Rnw
eisa: some functions were renamed, and we don't use generics any more.

The simplest way to run ISA is to choose the two threshold parameters
and then call the \Rfunction{isa} function on the
\Rclass{ExpressionSet} object:
<<simple ISA>>=
thr.gene <- 2.7
thr.cond <- 1.4
set.seed(1) # to get the same results, always
modules <- isa(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
@

This first applies a non-specific filter to the data set filters the
data set and then runs ISA from
\Sexpr{formals(eisa:::isa.ExpressionSet)$no.seeds} % $
random seeds (the default). See Section~\ref{sec:detailed-isa} if the
default parameters are not appropriate for you and need more control.

@@

To perform enrichment analysis against the Gene Ontology database, all
you have to do is to supply your \Rclass{ISAModules} object to the
\Rfunction{isa.GO} function.
<<GO enrichment>>=
GO <- isa.GO(modules)
@

The \Rfunction{isa.GO} requires an annotation package that maps the
Entrez identifiers of the studied organism to GO terms. For human, the
\Rpackage{org.Hs.eg.db} package is required, for the mouse the
\Rpackage{org.Mm.eg.db} package, etc.

@@

For extracting the results themselves, the \Rfunction{summary}
function can be used, this converts them to a simple data frame. The
$p$-value limit can be supplied to \Rfunction{summary}. Note, that
since \Rfunction{isa.GO} calculates enrichment for many gene sets
(i.e. for all biclusters), \Rfunction{summary} returns a list of data
frames, one for each bicluster. A table for the first module:
<<GO summary>>=
summary(GO[[1]], p=0.001)[[2]][,-6]
@

@@

In addition, the following functions are implemented to work on the
objects returned by \Rfunction{isa.GO}: \Rfunction{htmlReport},
\Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
\Rfunction{exprectedCounts}, \Rfunction{universeCounts},
\Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
\Rfunction{geneIdUniverse}. These functions do essentially the same as
they counterparts for \Rclass{GOHyperGResult} objects, see the
documentation of the \Rpackage{GOstats} package. The only difference
is, that since here we are testing a list of gene sets (=biclusters),
they calculate the results for all gene sets and return a list.

@@

By default, the \Rfunction{isa.GO} function performs multiple testing
correction using the Holm method, this can be changed via the
\texttt{correction} and \texttt{correction.method} arguments. See the
manual page of the \Rfunction{p.adjust} function for the possible
multiple testing correction schemes.

@@

Enrichment calculation against the KEGG pathway goes essentially the
same way as for the Gene Ontology, this time we use the
\Rfunction{isa.KEGG} function:
<<KEGG enrichment>>=
KEGG <- isa.KEGG(modules)
KEGG
summary(KEGG)[[4]]
library(KEGG.db)
mget(sigCategories(KEGG)[[4]], KEGGPATHID2NAME)
@

@@

The \Rpackage{eisa} includes a simple way to check whether the genes
in a bicluster are associated with a chromosome. See the
\Rfunction{isa.CHR} function:
<<CHR enrichment>>=
CHR <- isa.CHR(modules)
summary(CHR,p=0.05)[[11]][,-6]
@

@@

The enrichment calculation itself is basically the same as for GO and
KEGG, but the \Rfunction{isa.miRNA} function should be used:
<<miRNA enrichment>>=
if (require(targetscan.Hs.eg.db)) {
  miRNA <- isa.miRNA(modules)
  summary(miRNA, p=0.05)[[11]]
}
@

@@

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{eisa.biclust} function converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
Bc <- eisa.biclust(modules)
Bc
@

@@

It is very possible that the ISA finds the same module more than once;
two or more seeds might converge to the same module. The function
\Rfunction{isa.unique} eliminates every module from the result of 
\Rfunction{isa.iterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

@@

It is possible to use non-random seeds as well. If you have some
knowledge about the data or are interested in a particular subset of
genes/samples, then you can feed in your seeds into the
\Rfunction{isa.iterate} function directly. In this case the
algorithm is deterministic, for the same seed you will always get the
same results. Using smart (i.e. non-random) seeds can be considered as
a semi-supervised approach.

@@

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{isa.normalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its genes (see the \Rfunction{scale} R function). $E_c$ is
calculated by centering and scaling the genes of $E$. $E_r$ is
used to calculate the sample signature of genes and $E_c$ is used
to calculate the signature of the samples.

It is possible to use another normalization. In this case the user is
requested to supply the normalized input data in a named list,
including the two matrices of appropriate
dimensions to the \Rfunction{isa.iterate} function.

@@

The two packages provide some functions that share the same name, and
the right one is selected based on the arguments supplied. E.g. both
\Rpackage{isa2} and \Rpackage{eisa} has a function called
\Rfunction{isa}. It the first argument of \Rfunction{isa} is an
\Rclass{ExpressionSet}, then the \Rpackage{eisa} version is called. If
the first argument is a matrix, then the \Rpackage{isa} version is
called.

This also means that there are two manual pages for the two
\Rfunction{isa} functions. I.e. if one types `\verb+?isa+', (and both
\Rpackage{isa2} and \Rpackage{eisa} are loaded, then one
has to choose between the two versions. To avoid this, the
\Rpackage{eisa} manual pages can also be referenced by supplying and
`\verb+e+' prefix: `\verb+?eisa+' or `\verb+?eisa.unique+'.

>>>>

The simplest way to run ISA is to choose the two threshold parameters
and then call the \Rfunction{isa} function on the
\Rclass{ExpressionSet} object:
<<simple ISA>>=
thr.gene <- 2.7
thr.cond <- 1.4
set.seed(1) # to get the same results, always
modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
@

This first applies a non-specific filter to the data set filters the
data set and then runs ISA from
\Sexpr{formals(eisa:::ISA)$no.seeds} % $
random seeds (the default). See Section~\ref{sec:detailed-isa} if the
default parameters are not appropriate for you and need more control.

@@

To perform enrichment analysis against the Gene Ontology database, all
you have to do is to supply your \Rclass{ISAModules} object to the
\Rfunction{ISA.GO} function.
<<GO enrichment>>=
GO <- ISA.GO(modules)
@

The \Rfunction{ISA.GO} requires an annotation package that maps the
Entrez identifiers of the studied organism to GO terms. For human, the
\Rpackage{org.Hs.eg.db} package is required, for the mouse the
\Rpackage{org.Mm.eg.db} package, etc.

@@

For extracting the results themselves, the \Rfunction{summary}
function can be used, this converts them to a simple data frame. The
$p$-value limit can be supplied to \Rfunction{summary}. Note, that
since \Rfunction{ISA.GO} calculates enrichment for many gene sets
(i.e. for all biclusters), \Rfunction{summary} returns a list of data
frames, one for each bicluster. A table for the first module:
<<GO summary>>=
summary(GO[[1]], p=0.001)[[2]][,-6]
@

@@

In addition, the following functions are implemented to work on the
objects returned by \Rfunction{ISA.GO}: \Rfunction{htmlReport},
\Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
\Rfunction{exprectedCounts}, \Rfunction{universeCounts},
\Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
\Rfunction{geneIdUniverse}. These functions do essentially the same as
they counterparts for \Rclass{GOHyperGResult} objects, see the
documentation of the \Rpackage{GOstats} package. The only difference
is, that since here we are testing a list of gene sets (=biclusters),
they calculate the results for all gene sets and return a list.

@@

By default, the \Rfunction{ISA.GO} function performs multiple testing
correction using the Holm method, this can be changed via the
\texttt{correction} and \texttt{correction.method} arguments. See the
manual page of the \Rfunction{p.adjust} function for the possible
multiple testing correction schemes.

@@

Enrichment calculation against the KEGG pathway goes essentially the
same way as for the Gene Ontology, this time we use the
\Rfunction{ISA.KEGG} function:
<<KEGG enrichment>>=
KEGG <- ISA.KEGG(modules)
KEGG
summary(KEGG)[[4]]
library(KEGG.db)
mget(sigCategories(KEGG)[[4]], KEGGPATHID2NAME)
@

@@

The \Rpackage{eisa} includes a simple way to check whether the genes
in a bicluster are associated with a chromosome. See the
\Rfunction{ISA.CHR} function:
<<CHR enrichment>>=
CHR <- ISA.CHR(modules)
summary(CHR,p=0.05)[[11]][,-6]
@

@@

The enrichment calculation itself is basically the same as for GO and
KEGG, but the \Rfunction{ISA.miRNA} function should be used:
<<miRNA enrichment>>=
if (require(targetscan.Hs.eg.db)) {
  miRNA <- ISA.miRNA(modules)
  summary(miRNA, p=0.05)[[11]]
}
@

@@

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{ISA.biclust} function converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
Bc <- ISA.biclust(modules)
Bc
@

@@

It is very possible that the ISA finds the same module more than once;
two or more seeds might converge to the same module. The function
\Rfunction{ISA.unique} eliminates every module from the result of 
\Rfunction{ISA.iterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

@@

It is possible to use non-random seeds as well. If you have some
knowledge about the data or are interested in a particular subset of
genes/samples, then you can feed in your seeds into the
\Rfunction{ISA.iterate} function directly. In this case the
algorithm is deterministic, for the same seed you will always get the
same results. Using smart (i.e. non-random) seeds can be considered as
a semi-supervised approach.

@@

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{ISA.normalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its genes (see the \Rfunction{scale} R function). $E_c$ is
calculated by centering and scaling the genes of $E$. $E_r$ is
used to calculate the sample signature of genes and $E_c$ is used
to calculate the signature of the samples.

It is possible to use another normalization. In this case the user is
requested to supply the normalized input data in a named list,
including the two matrices of appropriate
dimensions to the \Rfunction{ISA.iterate} function.

@@
